---
title: "Reproducible Data Programing Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Load and preprocess the data

```{r }
temp <-tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)
actdata<-read.csv(unzip(temp),  header = TRUE)
unlink(temp)
rm(temp)
```
 Take a quick look at the data
```{r}
str(actdata)

```
 Convert 'date' column to 'Date' type

```{r}
actdata$date<- as.Date(actdata$date, '%Y-%m-%d')
```
 Check the distribution of 'na' accross the data 
```{r}
naind<-apply(actdata,2, is.na)
natotal <-apply(naind,2,sum)
natotal
rm(naind)
```


# 2. Caclulate mean total number of steps taken per day

```{r message=FALSE, warning=FALSE}
library(dplyr)

dailyactivity <-summarize(group_by(actdata, date),totalsteps = sum(steps, na.rm = T) )
```


### Make a histogram of the total number of steps taken each day

```{r}
par(cex.axis = 0.7)
barplot(dailyactivity$totalsteps, names.arg =format(dailyactivity$date,format = "%m-%d"), main = 'Daily Steps Original Dataset', las=2,  xlab = "Date", ylab = 'Steps' )
dev.off()
```

### Calculate and report the mean and median total number of steps taken per day

```{r}
dailysummary<-c(mean = mean(dailyactivity$totalsteps, na.rm = T), median =median(dailyactivity$totalsteps, na.rm = T))
dailysummary
```


# 3. Analyze the average daily activity pattern

### Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days 
```{r}
pattern<-summarize(group_by(actdata, interval),averagesteps = mean(steps, na.rm = T) )
plot(pattern$interval, pattern$averagesteps, type = 'l', main = "Diurnal pattern", ylab = "Average steps", xlab = "Interval number")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxinterval <- pattern$interval[ind =which(pattern$averagesteps ==max(pattern$averagesteps))]
maxinterval
```


# 4. Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
see part 1, variable 'natotal'
```{r}
natotal
```

### Devise a strategy for filling in all of the missing values in the dataset.
Missing step values will be replaced with mean steps for corresponing 5-min interval

### Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
naninterval <-actdata$interval[ is.na(actdata$steps)]

stepsimp <- vector('numeric',length(naninterval))
for (i in 1:length(naninterval)){
  stepsimp[i]<-pattern$averagesteps[pattern$interval == naninterval[i]]
  
}

actdataimp <- actdata
actdataimp$steps[is.na(actdata$steps)]<-stepsimp
```
### Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 

```{r}
dailyactimp <-summarize(group_by(actdataimp, date),totalsteps = sum(steps) )

par(cex.axis = 0.7)
barplot(dailyactimp$totalsteps, names.arg =format(dailyactimp$date,format = "%m-%d"), main = 'Daily Steps Imputed ', las=2,  xlab = "Date", ylab = 'Steps'  )
dev.off()

dailysumimp<-c(mean = mean(dailyactimp$totalsteps), median =median(dailyactimp$totalsteps))
dailysumimp
```
### Do these values differ from the estimates from the first part of the assignment?What is the impact of imputing missing data on the estimates of the total daily number of steps?

Imputing missing data resulted in a modest (~ 15% [10766/9354]) increase in the estimate of mean daily steps, and almost did not change [10766/10395] the estimate of the median daily steps 



# 5. The differences in activity patterns between weekdays and weekends

### Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
factorday<- factor(weekdays(actdataimp$date), levels =c('Monday','Tuesday', 'Wednesday',
                                                    'Thursday', 'Friday','Saturday','Sunday'), ordered = T)

levels(factorday)[1:5]<- 'weekday'
levels(factorday)[2:3]<- 'weekend'

actdataimp$factorday<-factorday

```
### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r fig,  fig.width = 5, fig.height= 7, fig.align = "center", warning=FALSE}
weekdayact <- summarize(actdataimp %>% group_by(interval, factorday),meansteps = mean(steps))

library(lattice)
xyplot(meansteps~ interval  | factorday, data = weekdayact, type='l')

```